---
title: "Mixed-Effects Model Analysis"
author: "Clara Rossignol"
date: "`r Sys.Date()`"
output: pdf_document
---

# **Introduction**

This analysis investigates the percentage of predicted dying neurons across different brain regions while comparing Alzheimer's disease (AD) vs. Control (CTRL) groups.\
A **linear mixed-effects model (LMM)** is used to account for region-specific variability.

------------------------------------------------------------------------

# **Load Required Libraries**

```{r setup, message=FALSE, warning=FALSE}
# Load necessary packages
library(readxl)      # Read Excel data
library(lme4)        # Mixed-effects modeling
library(lmerTest)    # p-values for lme4 models
library(ggplot2)     # Data visualization
library(dplyr)       # Data wrangling
library(effsize)     # Effect size (Cohen's d)
```

# **Load and Inspect Data**

```{r}
# Define the data file path (adjust if necessary)
data_file <- "E:/Clara/Mixed-effects_Model/Processed_DyingNeurons_Data.xlsx"

# Read the data
df <- read_excel(data_file)

# Display first few rows of the dataset
head(df)
```

# **Preprocessing**

```{r}
# Convert categorical variables to factors
df$Region <- as.factor(df$Region)
df$Group <- as.factor(df$Group)

# Ensure consistent naming (CTRL → Ctrl)
df$Group <- recode_factor(df$Group, "CTRL" = "Ctrl", "AD" = "AD")

# Summary of dataset
summary(df)
```

# **Visualizing the Data**

```{r}
# Convert PercentageDyingNeurons to numeric if it's not already
df$PercentageDyingNeurons <- as.numeric(df$PercentageDyingNeurons)

# Boxplot of Percentage of Dying Neurons by Group and Region
ggplot(df, aes(x = Group, y = PercentageDyingNeurons, fill = Group)) +
  geom_boxplot(alpha = 0.5, width = 0.6, outlier.shape = NA) +  # Transparent boxplots
  geom_jitter(position = position_jitter(width = 0.2, height = 0), size = 2, alpha = 0.7) +  # Jittered points
  facet_wrap(~ Region) +  # Separate by brain region
  theme_minimal() +  # Clean theme
  scale_fill_manual(values = c("Ctrl" = "#E69F00", "AD" = "#56B4E9")) +  # Custom colors
  scale_y_continuous(breaks = pretty(df$PercentageDyingNeurons, n = 5)) +  # Reduce y-axis labels
  labs(
    title = "Comparison of Percentage of Dying Neurons",
    x = "Group",
    y = "Percentage of Dying Neurons"
  ) +
  theme(
    legend.position = "right",
    text = element_text(size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# **Fitting the Mixed-Effects Model**

```{r}
# Define the mixed-effects model:
# - Fixed effect: Group (AD vs. Ctrl)
# - Random effect: Region (to account for region-specific variability)
model <- lmer(PercentageDyingNeurons ~ Group + (1 | Region), data = df)

# Display model summary
summary(model)

confint(model)
```
# **Effect Size Calculation (Cohen's d)**

```{r}
# Compute Cohen's d for AD vs. CTRL
cohens_d <- cohen.d(df$PercentageDyingNeurons[df$Group == "AD"], 
                    df$PercentageDyingNeurons[df$Group == "Ctrl"])
print(cohens_d)
```

A Cohen's d of:

- 0.2 = Small effect 
- 0.5 = Medium effect
- 0.8+ = Large effect


# **Checking Model Assumptions**

```{r}
# Residuals plot to check homoscedasticity
plot(model)

# Q-Q plot for normality of residuals
qqnorm(resid(model))
qqline(resid(model))

# Histogram of residuals to check distribution shape
hist(resid(model), breaks = 20, main = "Residuals Distribution", xlab = "Residuals")
```

**Model Assumptions Summary**

- Homoscedasticity: Residuals appear evenly spread, no major funneling.
- Normality: Q-Q plot suggests residuals are normally distributed, minor deviations at the tails.
- Residual Distribution: Slight skew, but overall normality is reasonable.

Since normality assumptions look acceptable on the Q-Q plot, we do not apply a log transformation.


